An Image-based method for phase estimation, gating and temporal super-resolution of \\cardiac ultrasound

### 2017-02-17 paper155.tex
Updated paper - finished phase estimation section

\begin{abstract}
Point-of-care low-cost hand-held ultrasound devices are emerging as an increasingly effective tool for rapid non-invasive visual assessment of cardiac structure and function on the bed-side. Their diagnostic value can be enriched further by coupling them with automated image analysis algorithms to accurately detect and quantify any cardiac abnormalities at the point-of-care potentially even in asymptomatic patients. However, a major challenge to this is the poor spatio-temporal resolution and high noise level of images obtained using these devices. In an effort to address this problem, we present a novel method for de-noising, gating, and temporal super-resolution of noisy low frame-rate periodic image sequences such as the ones obtained using low-cost hand-held cardiac ultrasound devices. We first leverage the inherent image-level redundancy in a periodic image sequence to decouple signal from noise using a low-rank + sparse decomposition algorithm. Next, we transform the complex high-dimensional periodic image sequence to a narrow-band 1D time series with the same periodicity characteristics using a novel strategy based on inter-frame similarity. Next, we decompose this time series use a season-trend decomposition method to obtain the heart-beat (season) and respiratory (trend) signals. Next, we use the Hilbert transform to estimate the instantaneous cardiac and respiratory phases of each frame to facilitate gating and super-resolution. Lastly, we use Nadarya-Watson kernel regression to construct an image manifold parameterized by intra-period phase which can then be sampled to reconstruct a de-noised single-period image sequence at a higher temporal-resolution. We demonstrate the utility and efficacy of our method both on synthetic data and on real image sequences of the beating heart acquired using a low-cost hand-held ultrasound device.
%\keywords{Cardiac, Ultrasound, Phase estimation, Gating, Temporal super-resolution}
\end{abstract}

@@

%\vspace{-0.5cm}
\section{Method}
\label{sec:method}
In this section, we present the theory underlying the proposed methods for instantaneous phase estimation, gating and temporal super-resolution of cardiac ultrasound videos. Specifically, we begin by describing our method for the estimation of instantaneous cardiac and respiratory phases in Section \ref{sec:method:phase_estimation}. We then present a robust method that uses these phase estimates to gate out video frames predominantly influenced by respiratory motion in Section \ref{sec:method:gating}. Finally, in Section \ref{sec:method:super_resolution}, we present a kernel regression method that uses the phase-tagged image sequence to reconstruct a single-period video at higher-temporal resolution. 
\begin{IonFigT}
\centering
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/ecg_instaphase_overlay.png}\\
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/qrs_peak_to_peak.png}
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/instaphase_valley_to_valley.png}
\ionbox{5.0in}\\
\caption{Illustration of correspondence between ECG and the instantaneous cardiac phase estimated using the proposed method: (a) ECG signal (blue) and the cardiac phase estimated using our method using only image data. Also shown are the peaks of the QRS complex (pink circle) and the minima of the estimated cardiac phase (black circle) that correspond to the QRS peaks, (b) Five video frames evenly spaced in time between the first and second QRS peaks of the ECG signal constituting one cardiac cycle, and (c) Five video frames evenly spaced in time between first and second minima of the estimated cardiac phase constituting one cardiac cycle.}
\label{fig:instaphase_vs_ecg}
\end{IonFigT}
%\vspace{-0.5cm}
\subsection{Estimation of instantaneous cardiac and respiratory phases}
\label{sec:method:phase_estimation}
\begin{IonFigT}
\centering
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/simMat.png}
\ionbox{4.75in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/season_trend_decomposition.png}
\ionbox{4.75in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/periodogram.png}
\ionbox{4.75in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/instantaneous_phase.png}
\ionbox{4.75in}\\
\caption{Illustration of the proposed phase estimation method: (a-left) Inter-frame similarity matrix, (a-mid) Trend component corresponding to the respiratory motion, (a-right) Seasonal component corresponding to the beating heart motion, (b) Similarity profile with minimum spectral entropy along with the associated trend/respiration and seasonal/heart-beat components, (c) Periodogram of the heart-beat and respiration components, and (d) Instantaneous phase of heart-beat and respiration components.}
\label{fig:phase_estimation}
\end{IonFigT}
While there have been numerous efforts for the estimation of instantaneous phase and/or frequency in periodic univariate time series data (wherein a single-variable is measured at each time point)\cite{Boashash1992,Lu2013,Luo2003}, we are not aware of any methods that tackle this problem in a multivariate setting which is the case with cardiac ultrasound videos wherein several thousands of variables (intensities of pixels) could be measured at each time point. Our strategy was to find a way to simplify this seemingly complex multi-variate problem by transforming it into a univariate one and take advantage of existing univariate methods to solve the problem at hand. To accomplish this, we compute the similarity/affinity between all pairs of frames in the given periodic image sequence containing $n$ images to create a matrix $A \in R^{n \times n}$ where in the element $A(i,j)$ is equal to the similarity between the $i^{th}$ and $j^{th}$ frames of the given sequence. Here in, we use normalization correlation also known as the Pearson correlation coefficient to quantify inter-frame similarity but, in principle, any of the similarity metrics used in image registration algorithms \cite{Goshtasby2012} can potentially be chosen. Each row in the inter-frame similarity matrix $A$ can now be seen as a univariate time series that measures the similarity of the corresponding frame in the image sequence with all other frames. Additionally, for a periodic image sequence, we can expect them to be periodic provided that the corresponding frame is not significantly corrupted by distortions such as noise and changes in global illumination over time among others.

To mitigate the potential problem of global illumintation changes over time, we detrend and extract the seasonal part of the time series corresponding to each row of $A$ by dividing it with the trend signal estimated using a locally-weighted regression algorithm called loess\cite{Cleveland1990}. Let $A_s$ be the matrix whose rows contain the seasonal part obtained by detrending the corresponding row in $A$. Next, we minimize the risk from noise, by finding the row $u(t)$ of $A_s$ whose power spectrum or periodogram has minimum entropy. In our experiments, the time series $u(t)$ was nearly sinusoidal in morphology with most of its energy in its power spectrum concentrated within a very narrow-band with the same periodicity characteristics as the image sequence it was derived from. Based on this observation, we decided to use its Hilbert transform \cite{Lu2013} to estimate the instantaneous intra-period phase  of each frame. Specifically, we compute the instantaneous phase $\phi(t) \in \left [  -\pi, \pi\right )$ of the time series $u(t)$ as follows:
\begin{equation}
\phi(t) = arctan \left( \frac{H_u(t)}{u(t)}\right)
\end{equation} 
where $H_u(t)$ is the Hilbert transform of $u(t)$. Lastly, we map the instantaneous phase $\phi(t)$ to the range $\left [  0, 1\right )$ which can then be used to obtain the intra-period phase of each frame.

%\vspace{-0.5cm}
\subsection{Gating out respiratory frames}
\label{sec:method:gating}
%
Once the instantaneous cardiac and respiratory phases of each frame has been estimated, it can be used to facilitate the extraction of quantitative measurements from a desired part/point of the periodic cycle, a process commonly referred to as gating. In this section, we present a robust method that uses these phase estimates to gate out video frames predominantly influenced by respiratory motion.
%
\begin{IonFigT}
\centering
%
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/respiratory_phase_gating.png}
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/robust_lowess_gating.png}
\ionbox{5.0in}\\
%
\caption{Illustration of our method for gating out respiratory frames: (a) (b)}
\label{fig:respiratory_gating}
\end{IonFigT}
%
%\vspace{-0.5cm}
\subsection{Reconstruction of one-period video at high temporal resolution}
\label{sec:method:super_resolution}
%
Once the instantaneous cardiac phase of each frame has been estimated, it can be used to enable spatio-temporal super-resolution by considering the set of images within each period of a periodic image sequence as distinct observations of the same data. In this section, we present a method that uses a phase-tagged periodic image sequence to reconstruct a single-period video at a higher temporal resolution.

>>>>

\begin{abstract}
Point-of-care low-cost hand-held ultrasound devices are emerging as an increasingly effective tool for rapid non-invasive visual assessment of cardiac structure and function on the bed-side. Their diagnostic value can be enriched further by coupling them with automated image analysis algorithms to accurately detect and quantify any cardiac abnormalities at the point-of-care potentially even in asymptomatic patients. However, a major challenge to this is the poor spatio-temporal resolution and high noise level of images obtained using these devices. In an effort to address this problem, we present a novel method for de-noising, gating, and temporal super-resolution of noisy low frame-rate periodic image sequences such as the ones obtained using low-cost hand-held cardiac ultrasound devices.

We first transform the complex high-dimensional periodic cardiac ultrasound image sequence to a narrow-band 1D time series with the same periodicity characteristics using a novel strategy based on inter-frame similarity. Next, we decompose this time series use a trend extraction technique to obtain the trend/respiration and residual/heart-beat component signals. Next, we use the Hilbert transform to estimate the instantaneous cardiac and respiratory phases of each frame to facilitate gating and temporal super-resolution. Lastly, we use kernel regression with a novel kernel to learn an image manifold parametrized by instantaneous cardiac phase that can be sampled to generate a single-cycle video at a higher temporal-resolution. We demonstrate and validate our methods using cardiac ultrasound videos and associated ECG recordings of 6 mice.
%\keywords{Cardiac, Ultrasound, Phase estimation, Gating, Temporal super-resolution}
\end{abstract}

@@

%\vspace{-0.5cm}
\section{Method}
\label{sec:method}
%
In this section, we present the theory underlying the proposed methods for instantaneous phase estimation, gating and temporal super-resolution of cardiac ultrasound videos. Specifically, we begin by describing our method for the estimation of instantaneous cardiac and respiratory phases in Section \ref{sec:method:phase_estimation}. We then present a robust method that uses these phase estimates to gate out video frames predominantly influenced by respiratory motion in Section \ref{sec:method:gating}. Finally, in Section \ref{sec:method:super_resolution}, we present a kernel regression method that uses the phase-tagged image sequence to reconstruct a single-period video at higher-temporal resolution. 
%
%
\begin{IonFigT}
\centering
%
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/ecg_instaphase_overlay.png}\\
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/qrs_peak_to_peak.png}
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/instaphase_valley_to_valley.png}
\ionbox{5.0in}\\
%
\caption{Illustration of correspondence between ECG and the instantaneous cardiac phase estimated using the proposed method: (a) ECG signal (blue) and the cardiac phase estimated using our method using only image data. Also shown are the peaks of the QRS complex (pink circle) and the minima of the estimated cardiac phase (black circle) that correspond to the QRS peaks, (b) Five video frames evenly spaced in time between the first and second QRS peaks of the ECG signal constituting one cardiac cycle, and (c) Five video frames evenly spaced in time between first and second minima of the estimated cardiac phase constituting one cardiac cycle.}
\label{fig:instaphase_vs_ecg}
\end{IonFigT}
%\vspace{-0.5cm}
\subsection{Estimation of instantaneous cardiac and respiratory phases}
\label{sec:method:phase_estimation}
%
\begin{IonFigT}
\centering
%
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/simMat.png}
\ionbox{1.6in}\ionbox{1.6in}\ionbox{1.6in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/season_trend_decomposition.png}
\ionbox{4.75in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/periodogram.png}
\ionbox{4.75in}\\
\includegraphics[width=4.75in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/instantaneous_phase.png}
\ionbox{4.75in}\\
%
\caption{Illustration of the proposed phase estimation method: (a) Inter-frame similarity matrix, (b) Trend component corresponding to the respiratory motion, (c) Seasonal component corresponding to the beating heart motion, (d) The similarity profile chosen for phase estimation along with the associated trend/respiration and seasonal/heart-beat components, (e) Periodogram of the heart-beat and respiration components along with the periodicity characteristics such as frequency and cycle duration, and (f) Instantaneous phase of the heart-beat and respiration components.}
\label{fig:phase_estimation}
\end{IonFigT}
%
While there have been numerous efforts for the estimation of instantaneous phase and/or frequency in periodic univariate time series data (wherein a single-variable is measured at each time point)\cite{Boashash1992,Lu2013,Luo2003}, there are not many methods that tackle this problem in a multivariate setting which is the case with periodic cardiac ultrasound videos wherein several thousands of variables (intensities of pixels) could be measured at each time point. Our strategy was to find a way to simplify this seemingly complex multi-variate problem by transforming it into a univariate one and take advantage of existing univariate methods to solve the problem at hand.

	We first compute the similarity between all pairs of frames in the given periodic image sequence containing $N$ images/frames to create a matrix $A \in R^{N \times N}$ where in the element $A(i,j)$ is equal to the similarity between the $i^{th}$ and $j^{th}$ frames of the given sequence. Here in, we use normalization correlation also known as the Pearson correlation coefficient to quantify inter-frame similarity but, in principle, any of the similarity metrics used in image registration algorithms \cite{Goshtasby2012} can potentially be chosen. Each row in the inter-frame similarity matrix $A$ can now be seen as a univariate time series that measures the similarity of the corresponding frame in the image sequence with all other frames.
If the similarity metric is chosen with care and if the corresponding frame is not significantly corrupted we can expect this time series to encompass the
predominant periodicity characteristics of the original image sequence. Two kinds of periodic motion are present in cardiac ultrasound videos, one corresponding to the beating heart motion and the other corresponding to respiratory motion. Figure \ref{fig:phase_estimation}(a) shows the inter-frame similarity matrix of one of our cardiac ultrasound videos computed using the normalized correlation metric wherein these two sources of periodicity can be clearly observed in the frame similarity signal of many rows/columns.

	Next, we use a trend extraction technique~\cite{Alexandrov2012} called the Hodrick-Prescott (HP) filter~\cite{Hodrick1997} to decompose the frame similarity signal $u(t)$ into a sum of two components: (i) Trend component $\tau(t)$ encompassing the periodicity characteristic of only the lower-frequency respiratory motion, and (ii) Residual component $r(t)$ encompassing the periodicity characteristic of only the higher-frequency beating heart motion. The HP filter, that is widely used in the field of econometrics, performs the decomposition of $u(t) = \tau(t) + r(t)$ by solving the following optimization problem:
\begin{equation}	
\argmin{\tau(t)} \left[ \sum_{t=1}^{N}  \left(u(t) - \tau(t) \right)^2  + \lambda \sum_{t=1}^{N-1} \left( \nabla^2 \tau(t) \right)^2 \right] 
\end{equation}
where $\nabla^2\tau(t) = \tau(t+1) - 2 \tau(t) + \tau(t-1)$ is the second-order difference or derivative that penalizes curvature of the trend signal. We use a ridge-regression implementation of the Hodrick-prescott filter~\cite{Yamada2015} with $\lambda=6400$. Let $A_r$ and $A_h$ be the matrices whose rows contain the trend/respiratory and the residual/heart-beat components, respectively, of the frame similarity signals in the corresponding rows of the similarity matrix $A$. Figures~\ref{fig:phase_estimation}(b,c) show the trend ($A_r$) and residual ($A_h$) matrices for one of our cardiac ultrasound videos. To minimize the effect of any noise on phase estimation, we pick frame similarity signal corresponding to the row of $A_h$ whose periodogram or power-frequency distribution has minimum entropy. Figure~\ref{fig:phase_estimation}(d) shows the selected frame similarity signal (blue) along with the associated trend/respiration (pink) and residual/heart-beat (green) components. Figure~\ref{fig:phase_estimation}(e) shows the periodogram of these two components. Through Figures~\ref{fig:phase_estimation}(b-e), notice how well the two periodic sources have been decoupled from the raw frame similarity signal. Also, in Figure~\ref{fig:phase_estimation}, notice that most of the energy in the periodogram or power-frequency distribution of the respiration and heart-beat components is concentrated within a narrow frequency band. Furthermore, notice how these 1D signals capture the periodicity characteristics (e.g. frequency in beats-per-minute (bpm), period or cycle duration in frames, number of cycles) of the respective sources of motion in the input high-dimensional periodic image sequence.

	Next, considering the narrow-band nature of the derived trend/respiratory and residual/heart-beat signals, we can use the Hilbert transform~\cite{Lu2013} to estimate the instantaneous intra-period phase of each frame. Specifically, we compute the instantaneous phase $\phi(t) \in \left [  -\pi, \pi\right )$ of the time series $u(t)$ as follows:
\begin{equation}
\phi(t) = arctan \left( \frac{H_u(t)}{u(t)}\right)
\end{equation} 
where $H_u(t)$ is the Hilbert transform of $u(t)$. We map $\phi(t)$ from range $\left [  -\pi, \pi\right )$ to $\left [  0, 1\right )$. Let $\phi_c(t)$ and $\phi_r(t)$ denote the instantaneous heart-beat/cardiac and respiratory phases computed from the trend/respiration and residual/heart-beat signals, respectively, using the aforementioned approach. Figure~\ref{fig:phase_estimation}(f) shows the instantaneous cardiac and respiratory phases computed from the heart-beat and respiration signals in in Figure~\ref{fig:phase_estimation}(d).

%\vspace{-0.5cm}
\subsection{Gating out respiratory frames}
\label{sec:method:gating}
%
Once the instantaneous cardiac and respiratory phases of each frame has been estimated, it can be used to facilitate the extraction of quantitative measurements from any desired part/point of the periodic cycle, a process commonly referred to as gating. In this section, we present a robust method that uses these phase estimates to gate out video frames predominantly influenced by respiratory motion.

	Firstly, based on the observation that the predominant influence of respiratory motion occurs around the troughs of the trend/respiration component (pink curve in Fig.~\ref{fig:phase_estimation}(d)), we identify the corresponding respiratory phase (equal to phase zero) and perform a rough initial gating by discarding the frames whose respiratory phase distance from it is below a specified cutoff value (we use 0.2). Figure~\ref{fig:respiratory_gating}(a) shows the discarded frames (blue circles) overlaid on the trend/respiration component, residual/heart-beat component, and instantaneous respiratory phase signals.

%
\begin{IonFigT}
\centering
%
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/respiratory_phase_gating.png}
\ionbox{5.0in}\\
\includegraphics[width=5.0in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/robust_lowess_gating.png}
\ionbox{5.0in}\\
%
\caption{Illustration of our method for gating out respiratory frames: (a) (b)}
\label{fig:respiratory_gating}
\end{IonFigT}
%
%\vspace{-0.5cm}
\subsection{Build model to generate image at any phase for super-resolution}
\label{sec:method:super_resolution}
%
Once the instantaneous cardiac phase of each frame has been estimated, it can be used to enable spatio-temporal super-resolution by considering the set of images within each period of a periodic image sequence as distinct observations of the same data. In this section, we present a method that uses a phase-tagged periodic image sequence to reconstruct a single-period video at a higher temporal resolution.
